---
title: "Baseline Models for Prediction"
author: "Ujjayini Das"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load `pacman` for pkg management
if (!require("pacman", character.only = TRUE)){
  install.packages("pacman", dep = TRUE)
  if (!require("pacman", character.only = TRUE))
    stop("Package not found")
}

# Packages to use
pkgs <- c("tidyverse",    # General data manipulation
          "dplyr",        # General data manipulation
          "dbplyr",       # Interface for databases
          "haven",        # Load Stata datasets
          "RSQLite",      # SQL
          "glue",         # For gluing
          "survey",       # Survey data analysis
          "srvyr",        # Survey data analysis
          "readxl",       # excel file reading
          "readr",        # reading text data
          "data.table",   # handling data table
          "purrr",
          "stringr",      # data wrangling for text
          "caret",
          "xgboost",
          "MLmetrics",
          "rpart",        # CART decision tree
          "randomForest", # Random forest
          "ROSE",          # Random oversampling examples - fix class imbalance
          "ggplot2",
          "kableExtra"
          )

pacman::p_load(char = pkgs)
rm(pkgs)
```

## Load data and preprocess for next steps

```{r}
# ------------------ Load all four raw wave files ----------------------------
w1 <- read.csv("Data/wave1_processed_withINCOME.csv")
w2 <- read.csv("Data/wave2_processed_withINCOME.csv")
w3 <- read.csv("Data/wave3_processed_withINCOME.csv")
w4 <- read.csv("Data/wave4_processed_withINCOME.csv")

# ---------- Load income files ------------
inc_w1 <- read.csv("Data/Wave1_raw_with_income.csv") %>%
  select(unique_ID, THTOTINC) %>%
  rename(ID = unique_ID, income_raw = THTOTINC)

inc_w2 <- read.csv("Data/Wave2_raw_with_income.csv") %>%
  select(unique_ID_2, THTOTINC) %>%
  rename(ID = unique_ID_2, income_raw = THTOTINC)

inc_w3 <- read.csv("Data/Wave3_raw_with_income.csv") %>%
  select(unique_ID_2, THTOTINC) %>%
  rename(ID = unique_ID_2, income_raw = THTOTINC)

inc_w4 <- read.csv("Data/Wave4_raw_with_income.csv") %>%
  select(unique_ID_2, THTOTINC) %>%
  rename(ID = unique_ID_2, income_raw = THTOTINC)

# Defining income levels for more variation
bin_income <- function(x) {
  case_when(
    is.na(x)       ~ "No income or loss",
    x <= 0         ~ "No income or loss",
    x < 1000       ~ "Less than $1,000",
    x < 2000       ~ "$1,000 to $1,999",
    x < 3000       ~ "$2,000 to $2,999",
    x < 5000       ~ "$3,000 to $4,999",
    x < 10000      ~ "$5,000 to $9,999",
    x < 25000      ~ "$10,000 to $24,999",
    x < 50000      ~ "$25,000 to $49,999",
    x < 75000      ~ "$50,000 to $74,999",
    TRUE           ~ "$75,000 or more"
  )
}

# Standardize the ID column name — Wave 1 uses unique_ID, others use unique_ID_2
# Strip leading zeros from Wave 1 IDs to match later waves
# Strip leading zeros from Wave 2-4 IDs to match Wave 1 format
strip_zeros <- function(id) sub("^0+", "", id)

# Join income into each processed wave file
w1 <- w1 %>%
  mutate(ID_nozero = strip_zeros(unique_ID))%>%
  left_join(inc_w1, by = c("ID_nozero" = "ID")) %>%
  mutate(income = bin_income(income_raw)) %>%
  select(-income_raw)

w2 <- w2 %>%
  mutate(ID_nozero = strip_zeros(unique_ID_2)) %>%
  left_join(inc_w2 %>% mutate(ID = strip_zeros(ID)), by = c("ID_nozero" = "ID")) %>%
  mutate(income = bin_income(income_raw)) %>%
  select(-income_raw, -unique_ID_2)

w3 <- w3 %>%
  mutate(ID_nozero = strip_zeros(unique_ID_2)) %>%
  left_join(inc_w3 %>% mutate(ID = strip_zeros(ID)), by = c("ID_nozero" = "ID")) %>%
  mutate(income = bin_income(income_raw)) %>%
  select(-income_raw, -unique_ID_2)

w4 <- w4 %>%
  mutate(ID_nozero = strip_zeros(unique_ID_2)) %>%
  left_join(inc_w4 %>% mutate(ID = strip_zeros(ID)), by = c("ID_nozero" = "ID")) %>%
  mutate(income = bin_income(income_raw)) %>%
  select(-income_raw, -unique_ID_2)
# verify income distribution
lapply(list(w1 = w1, w2 = w2, w3 = w3, w4 = w4), function(df) {
  table(df$income, useNA = "ifany")
})

# ---------------- Prepare datasets for longitudinal prediction -----------------------
# wave 1 data set - no longitudinal predictor
df_wave1 <- w1 %>%
  filter(imputed_snap == FALSE) %>%
  select(ID_nozero, marital.status, sex, hispanic.origin, race,
         education, citizenship, employment, state, age, income, SNAP)

# wave 2 dataset - only wave 1 as longitudinal predictor
# Pull Wave 1 SNAP outcome for each person (non-imputed only)
snap_w1 <- w1 %>%
  filter(imputed_snap == FALSE) %>%
  select(ID_nozero, SNAP_wave1 = SNAP)

# Merge Wave 2 data with Wave 1 outcome
df_wave2 <- w2 %>%
  filter(imputed_snap == FALSE) %>%
  inner_join(snap_w1, by = "ID_nozero") %>%   # inner_join = only those in BOTH waves
  mutate(SNAP = ifelse(snap == "SNAP covered", "Yes", "No")) %>%
  select(ID_nozero, marital.status, sex, hispanic.origin, race,
         education, citizenship, employment, state, age, income,
         SNAP_wave1,   # <-- longitudinal predictor
         SNAP)         # <-- outcome to predict

# wave 3 dataset - wave 1 and 2 as longitudinal predictors
snap_w2 <- w2 %>%
  filter(imputed_snap == FALSE) %>%
  mutate(SNAP_wave2 = ifelse(snap == "SNAP covered", "Yes", "No")) %>%
  select(ID_nozero, SNAP_wave2)

df_wave3 <- w3 %>%
  filter(imputed_snap == FALSE) %>%
  inner_join(snap_w1, by = "ID_nozero") %>%
  inner_join(snap_w2, by = "ID_nozero") %>%
  mutate(SNAP = ifelse(snap == "SNAP covered", "Yes", "No")) %>%
  select(ID_nozero, marital.status, sex, hispanic.origin, race,
         education, citizenship, employment, state, age, income,
         SNAP_wave1, SNAP_wave2,   # <-- longitudinal predictors
         SNAP)

# wave 4 dataset - wave 1,2,3 as longitudinal predictors
snap_w3 <- w3 %>%
  filter(imputed_snap == FALSE) %>%
  mutate(SNAP_wave3 = ifelse(snap == "SNAP covered", "Yes", "No")) %>%
  select(ID_nozero, SNAP_wave3)

df_wave4 <- w4 %>%
  filter(imputed_snap == FALSE) %>%
  inner_join(snap_w1, by = "ID_nozero") %>%
  inner_join(snap_w2, by = "ID_nozero") %>%
  inner_join(snap_w3, by = "ID_nozero") %>%
  mutate(SNAP = ifelse(snap == "SNAP covered", "Yes", "No")) %>%
  select(ID_nozero, marital.status, sex, hispanic.origin, race,
         education, citizenship, employment, state, age, income,
         SNAP_wave1, SNAP_wave2, SNAP_wave3,   # <-- longitudinal predictors
         SNAP)

# Prepare train-test datasets
prepare_split <- function(df, seed, train_prop) {
  df_model <- df %>%
    select(-ID_nozero) %>%
    mutate(across(where(is.character), as.factor)) %>%  # convert all character to factor
    mutate(SNAP = factor(SNAP, levels = c("No", "Yes")))
  
  set.seed(seed)
  idx   <- createDataPartition(df_model$SNAP, p = train_prop, list = FALSE)
  train <- df_model[ idx, ]
  test  <- df_model[-idx, ]
  
  list(train = train, test = test)
}
split_w1 <- prepare_split(df_wave1, seed = 24, train_prop = 0.8)
split_w2 <- prepare_split(df_wave2, seed = 24, train_prop = 0.7)
split_w3 <- prepare_split(df_wave3, seed = 24, train_prop = 0.7)
split_w4 <- prepare_split(df_wave4, seed = 24, train_prop = 0.7)

splits <- list(
  w1 = split_w1,
  w2 = split_w2,
  w3 = split_w3,
  w4 = split_w4
)

# Check class distribution of SNAP on the training and test data
check_class_dist <- function(split, wave_name) {
  bind_rows(
    split$train %>% mutate(set = "train"),
    split$test  %>% mutate(set = "test")
  ) %>%
    count(set, SNAP) %>%
    group_by(set) %>%
    mutate(pct = round(n / sum(n) * 100, 1)) %>%
    mutate(wave = wave_name)
}

bind_rows(
  check_class_dist(splits$w1, "Wave 1"),
  check_class_dist(splits$w2, "Wave 2"),
  check_class_dist(splits$w3, "Wave 3"),
  check_class_dist(splits$w4, "Wave 4")
) %>% print(n = 50)
## Found high class imbalance

## Check missingness
check_missingness <- function(split, wave_name) {
  total_na <- sum(is.na(split$train)) + sum(is.na(split$test))
  cat(wave_name, "— total NA values:", total_na, "\n")
  if (total_na > 0) {
    cat("  Columns with NA in train:\n")
    print(colSums(is.na(split$train))[colSums(is.na(split$train)) > 0])
  }
}

check_missingness(splits$w1, "Wave 1")
check_missingness(splits$w2, "Wave 2")
check_missingness(splits$w3, "Wave 3")
check_missingness(splits$w4, "Wave 4")

## found marital status to be missing in wave 3 and 4 for small number of cases (1%) but I am not deleting those cases or imputing them. Rather, I am using 'unknown' as a separate category for these to ensure similar treatment as it had with LLMs.
# Apply to all four splits — replace NA in marital.status with "Unknown"
add_unknown_marital <- function(split) {
  fix <- function(df) {
    df %>%
      mutate(marital.status = as.character(marital.status),
             marital.status = replace_na(marital.status, "Unknown"),
             marital.status = factor(marital.status))
  }
  list(train = fix(split$train),
       test  = fix(split$test))
}

splits <- lapply(splits, add_unknown_marital)

# Save all four wave splits as CSVs for Python to read tomorrow
for (wave in c("w1", "w2", "w3", "w4")) {
  write.csv(splits[[wave]]$train, 
            paste0("train_", wave, ".csv"), row.names = FALSE)
  write.csv(splits[[wave]]$test,  
            paste0("test_",  wave, ".csv"), row.names = FALSE)
}

```

## Run baseline models

I am choosing logistic regression, xgboost, CART, random forest and majority class classifier for comparison.
UPDATE: 4/18 - Reporting LR, RF and XGBoost only.

```{r}
# Function to compute Expected Calibration Error (ECE) from paper (pg. 9) for baseline models
compute_ece <- function(prob_yes, actuals, n_bins = 10) {
  # prob_yes : numeric vector of P(SNAP = Yes)
  # actuals  : factor with levels c("No", "Yes")
  
  y_bin      <- as.integer(actuals == "Yes")
  bin_edges  <- seq(0, 1, length.out = n_bins + 1)
  ece        <- 0.0
  n          <- length(prob_yes)
  
  for (i in seq_len(n_bins)) {
    in_bin <- prob_yes >= bin_edges[i] & prob_yes < bin_edges[i + 1]
    # Include upper edge in last bin
    if (i == n_bins) in_bin <- prob_yes >= bin_edges[i] & prob_yes <= bin_edges[i + 1]
    if (sum(in_bin) == 0) next
    
    bin_conf <- mean(prob_yes[in_bin])
    bin_acc  <- mean(y_bin[in_bin])
    ece      <- ece + (sum(in_bin) / n) * abs(bin_conf - bin_acc)
  }
  round(ece, 4)
}

# Function to evaluate the model performance and report metrics such as accuracy, recall, precision, F1 and ece
evaluate_model <- function(model_fit, test_df, outcome_col,
                           model_name, wave_name, n_bins = 10) {
  # model_fit  : trained caret model object
  # test_df    : test data frame (includes outcome_col)
  # outcome_col: character name of the SNAP column, e.g. "SNAP"

  actuals     <- test_df[[outcome_col]]
  predictions <- predict(model_fit, newdata = test_df, type = "raw")
  probs       <- predict(model_fit, newdata = test_df, type = "prob")[, "Yes"]

  cm <- confusionMatrix(predictions, actuals,
                        positive = "Yes", mode = "prec_recall")
  tbl <- cm$table

  TP <- tbl["Yes", "Yes"]; TN <- tbl["No",  "No"]
  FP <- tbl["Yes", "No"];  FN <- tbl["No",  "Yes"]

  precision_yes <- if ((TP + FP) > 0) TP / (TP + FP) else NA_real_
  recall_yes    <- if ((TP + FN) > 0) TP / (TP + FN) else NA_real_
  f1_yes        <- if (!is.na(precision_yes) & !is.na(recall_yes) &
                       (precision_yes + recall_yes) > 0)
                     2 * precision_yes * recall_yes /
                       (precision_yes + recall_yes) else NA_real_

  precision_no  <- if ((TN + FN) > 0) TN / (TN + FN) else NA_real_
  recall_no     <- if ((TN + FP) > 0) TN / (TN + FP) else NA_real_
  f1_no         <- if (!is.na(precision_no) & !is.na(recall_no) &
                       (precision_no + recall_no) > 0)
                     2 * precision_no * recall_no /
                       (precision_no + recall_no) else NA_real_

  accuracy <- (TP + TN) / (TP + TN + FP + FN)
  ece      <- compute_ece(probs, actuals, n_bins = n_bins)

  bind_rows(
    tibble(Wave = wave_name, Model = model_name, Class = "Yes",
           N_test    = length(actuals),
           Accuracy  = round(accuracy,      4),
           Precision = round(precision_yes, 4),
           Recall    = round(recall_yes,    4),
           F1        = round(f1_yes,        4),
           ECE       = ece),
    tibble(Wave = wave_name, Model = model_name, Class = "No",
           N_test    = length(actuals),
           Accuracy  = round(accuracy,     4),
           Precision = round(precision_no, 4),
           Recall    = round(recall_no,    4),
           F1        = round(f1_no,        4),
           ECE       = ece)          # ECE is model-level, same for both rows
  )
}

# ----- Majority class classifier ----------
### NOT INCLUDED ######
run_majority <- function(split, wave_name) {
  preds <- factor(rep("No", nrow(split$test)), levels = c("No", "Yes"))
  evaluate_model(preds, split$test$SNAP, "Majority Class", wave_name)
}

res_maj <- bind_rows(
  run_majority(splits$w1, "Wave 1"),
  run_majority(splits$w2, "Wave 2"),
  run_majority(splits$w3, "Wave 3"),
  run_majority(splits$w4, "Wave 4")
)

# ----- Logistic regression -----------
run_logistic <- function(split, wave_name) {
  ctrl  <- trainControl(method = "none", classProbs = TRUE)
  model <- train(SNAP ~ ., data = split$train,
                 method = "glm", family = "binomial",
                 trControl = ctrl)
  evaluate_model(model, split$test, "SNAP", "Logistic Regression", wave_name)
}

res_lr <- bind_rows(
  run_logistic(splits$w1, "Wave 1"),
  run_logistic(splits$w2, "Wave 2"),
  run_logistic(splits$w3, "Wave 3"),
  run_logistic(splits$w4, "Wave 4")
)

# ------------- XGBoost --------------------------
run_xgboost <- function(split, wave_name) {
  ctrl <- trainControl(method = "none",
                       classProbs    = TRUE,   
                       savePredictions = "final")
  
  # Must supply exactly one row when method = "none"
  xgb_grid <- expand.grid(
    nrounds          = 100,
    max_depth        = 6,
    eta              = 0.3,
    gamma            = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample        = 0.8
  )
  
  model <- train(SNAP ~ ., data = split$train,
                 method     = "xgbTree",
                 trControl  = ctrl,
                 tuneGrid   = xgb_grid,
                 verbosity  = 0)
  
  evaluate_model(model, split$test, "SNAP", "XGBoost", wave_name)
}

res_xgb <- bind_rows(
  run_xgboost(splits$w1, "Wave 1"),
  run_xgboost(splits$w2, "Wave 2"),
  run_xgboost(splits$w3, "Wave 3"),
  run_xgboost(splits$w4, "Wave 4")
)

## CART
run_cart <- function(split, wave_name) {
  ctrl  <- trainControl(method = "none", classProbs = TRUE)
  model <- train(SNAP ~ ., data = split$train,
                 method = "rpart",
                 trControl = ctrl)
  evaluate_model(model, split$test, "SNAP", "CART", wave_name)
}


res_cart <- bind_rows(
  run_cart(splits$w1, "Wave 1"),
  run_cart(splits$w2, "Wave 2"),
  run_cart(splits$w3, "Wave 3"),
  run_cart(splits$w4, "Wave 4")
)

## Random forest
run_rf <- function(split, wave_name) {
  ctrl  <- trainControl(method = "none", classProbs = TRUE)
  model <- train(SNAP ~ ., data = split$train,
                 method    = "rf",
                 trControl = ctrl,
                 tuneGrid  = data.frame(mtry = floor(sqrt(ncol(split$train) - 1))),
                 ntree     = 100)
  evaluate_model(model, split$test, "SNAP", "Random Forest", wave_name)
}
res_rf <- bind_rows(
  run_rf(splits$w1, "Wave 1"),
  run_rf(splits$w2, "Wave 2"),
  run_rf(splits$w3, "Wave 3"),
  run_rf(splits$w4, "Wave 4")
)

results_baselines <- bind_rows(
  #res_maj,
  res_lr,
  res_cart,
  res_rf,
  res_xgb
) %>%
  arrange(Wave, Model)

print(results_baselines, n = 60)
write.csv(results_baselines, "baseline_results.csv", row.names = FALSE)


```

## Handling class imbalance - minority class oversampled in training set

```{r}
# ── Balanced runner for caret-compatible models (LR, CART, RF) ───────────────
run_model_balanced <- function(split, wave_name, method, model_name,
                               tuneGrid = NULL) {
  train_bal <- ROSE(SNAP ~ ., data = split$train, seed = 22)$data %>%
    mutate(SNAP = factor(SNAP, levels = c("No", "Yes")))

  ctrl <- trainControl(method = "none", classProbs = TRUE)  
  args <- list(SNAP ~ ., data = train_bal,
               method = method, trControl = ctrl)
  if (!is.null(tuneGrid)) args$tuneGrid <- tuneGrid
  if (method == "glm")    args$family   <- "binomial"

  model <- do.call(train, args)

  
  evaluate_model(model, split$test, "SNAP", model_name, wave_name)
}

# ---- XGBoost with scale_pos_weight (native xgboost, not caret) -------------
run_xgboost_weighted <- function(split, wave_name) {
  X_train <- model.matrix(SNAP ~ . - 1, data = split$train)
  X_test  <- model.matrix(SNAP ~ . - 1, data = split$test)
  y_train <- as.numeric(split$train$SNAP == "Yes")
  y_test  <- split$test$SNAP

  spw    <- sum(y_train == 0) / sum(y_train == 1)
  dtrain <- xgb.DMatrix(X_train, label = y_train)
  dtest  <- xgb.DMatrix(X_test)

  params <- list(objective        = "binary:logistic",
                 eval_metric      = "logloss",
                 eta              = 0.1,
                 max_depth        = 6,
                 scale_pos_weight = spw,
                 nthread          = 4)

  model <- xgb.train(params, dtrain, nrounds = 100, verbose = 0)
  probs <- predict(model, dtest)                                    # P(SNAP = Yes)
  preds <- factor(ifelse(probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

  # ------- Manual evaluation (native xgboost has no caret interface) -----------------
  cm  <- confusionMatrix(preds, y_test, positive = "Yes", mode = "prec_recall")
  tbl <- cm$table
  TP  <- tbl["Yes","Yes"]; TN <- tbl["No","No"]
  FP  <- tbl["Yes","No"];  FN <- tbl["No","Yes"]

  prec_yes <- if ((TP+FP)>0) TP/(TP+FP) else NA_real_
  rec_yes  <- if ((TP+FN)>0) TP/(TP+FN) else NA_real_
  f1_yes   <- if (!is.na(prec_yes)&!is.na(rec_yes)&(prec_yes+rec_yes)>0)
                2*prec_yes*rec_yes/(prec_yes+rec_yes) else NA_real_
  prec_no  <- if ((TN+FN)>0) TN/(TN+FN) else NA_real_
  rec_no   <- if ((TN+FP)>0) TN/(TN+FP) else NA_real_
  f1_no    <- if (!is.na(prec_no)&!is.na(rec_no)&(prec_no+rec_no)>0)
                2*prec_no*rec_no/(prec_no+rec_no) else NA_real_
  acc      <- (TP+TN)/(TP+TN+FP+FN)
  ece      <- compute_ece(probs, y_test)

  bind_rows(
    tibble(Wave=wave_name, Model="XGBoost (Weighted)", Class="Yes",
           N_test=length(y_test), Accuracy=round(acc,4),
           Precision=round(prec_yes,4), Recall=round(rec_yes,4),
           F1=round(f1_yes,4), ECE=ece),
    tibble(Wave=wave_name, Model="XGBoost (Weighted)", Class="No",
           N_test=length(y_test), Accuracy=round(acc,4),
           Precision=round(prec_no,4), Recall=round(rec_no,4),
           F1=round(f1_no,4), ECE=ece)
  )
}

# ----------- Run all balanced models ----------------------------
wave_list  <- list(w1 = "Wave 1", w2 = "Wave 2", w3 = "Wave 3", w4 = "Wave 4")

res_balanced <- bind_rows(
  # Logistic Regression (ROSE)
  map_dfr(names(wave_list), ~run_model_balanced(
    splits[[.x]], wave_list[[.x]], "glm", "Logistic Regression (ROSE)")),

  # CART (ROSE)
  map_dfr(names(wave_list), ~run_model_balanced(
    splits[[.x]], wave_list[[.x]], "rpart", "CART (ROSE)",
    tuneGrid = data.frame(cp = 0.01))),

  # Random Forest (ROSE)
  map_dfr(names(wave_list), ~run_model_balanced(
    splits[[.x]], wave_list[[.x]], "rf", "Random Forest (ROSE)",
    tuneGrid = data.frame(mtry = floor(sqrt(ncol(splits[[.x]]$train) - 1))))),

  # XGBoost (scale_pos_weight — native)
  map_dfr(names(wave_list), ~run_xgboost_weighted(splits[[.x]], wave_list[[.x]]))
)

# ── Combine with baseline results ────────────────────────────────────────────
results_all <- bind_rows(results_baselines, res_balanced) %>%
  arrange(Wave, Model, Class)

write.csv(results_all, "baseline_results_full.csv", row.names = FALSE)

```

# Ablation experiment - temporal signal decomposition 

```{r}
## --- Ablation runner for caret-compatible models -------------------------
run_ablation <- function(split, wave_name, predictors, label, method,
                         tuneGrid = NULL) {
  df_tr <- split$train %>% select(all_of(c(predictors, "SNAP")))
  df_te <- split$test  %>% select(all_of(c(predictors, "SNAP")))

  ctrl <- trainControl(method = "none", classProbs = TRUE)  # classProbs for ECE
  args <- list(SNAP ~ ., data = df_tr, method = method, trControl = ctrl)
  if (method == "glm")    args$family   <- "binomial"
  if (!is.null(tuneGrid)) args$tuneGrid <- tuneGrid

  model <- do.call(train, args)

  
  evaluate_model(model, df_te, "SNAP", label, wave_name)
}

# ----------- XGBoost ablation — native xgboost, manual evaluation + ECE ----------------------
run_ablation_xgb <- function(split, wave_name, predictors, label) {
  df_tr <- split$train %>% select(all_of(c(predictors, "SNAP")))
  df_te <- split$test  %>% select(all_of(c(predictors, "SNAP")))

  X_tr  <- model.matrix(SNAP ~ . - 1, data = df_tr)
  X_te  <- model.matrix(SNAP ~ . - 1, data = df_te)
  y_tr  <- as.numeric(df_tr$SNAP == "Yes")
  y_te  <- df_te$SNAP

  model <- xgb.train(
    list(objective = "binary:logistic", eta = 0.1,
         max_depth = 6, nthread = 4),
    xgb.DMatrix(X_tr, label = y_tr),
    nrounds = 100, verbose = 0
  )

  probs <- predict(model, xgb.DMatrix(X_te))
  preds <- factor(ifelse(probs >= 0.5, "Yes", "No"), levels = c("No", "Yes"))

  
  cm  <- confusionMatrix(preds, y_te, positive = "Yes", mode = "prec_recall")
  tbl <- cm$table
  TP  <- tbl["Yes","Yes"]; TN <- tbl["No","No"]
  FP  <- tbl["Yes","No"];  FN <- tbl["No","Yes"]

  prec_yes <- if ((TP+FP)>0) TP/(TP+FP) else NA_real_
  rec_yes  <- if ((TP+FN)>0) TP/(TP+FN) else NA_real_
  f1_yes   <- if (!is.na(prec_yes)&!is.na(rec_yes)&(prec_yes+rec_yes)>0)
                2*prec_yes*rec_yes/(prec_yes+rec_yes) else NA_real_
  prec_no  <- if ((TN+FN)>0) TN/(TN+FN) else NA_real_
  rec_no   <- if ((TN+FP)>0) TN/(TN+FP) else NA_real_
  f1_no    <- if (!is.na(prec_no)&!is.na(rec_no)&(prec_no+rec_no)>0)
                2*prec_no*rec_no/(prec_no+rec_no) else NA_real_
  acc      <- (TP+TN)/(TP+TN+FP+FN)
  ece      <- compute_ece(probs, y_te)

  bind_rows(
    tibble(Wave=wave_name, Model=label, Class="Yes",
           N_test=length(y_te), Accuracy=round(acc,4),
           Precision=round(prec_yes,4), Recall=round(rec_yes,4),
           F1=round(f1_yes,4), ECE=ece),
    tibble(Wave=wave_name, Model=label, Class="No",
           N_test=length(y_te), Accuracy=round(acc,4),
           Precision=round(prec_no,4), Recall=round(rec_no,4),
           F1=round(f1_no,4), ECE=ece)
  )
}

# --------------- Predictor sets-------------------------------------------
demo_vars <- c("marital.status","sex","hispanic.origin","race",
               "education","citizenship","employment","state","age","income")
snap_w2   <- c("SNAP_wave1")
snap_w3   <- c("SNAP_wave1","SNAP_wave2")
snap_w4   <- c("SNAP_wave1","SNAP_wave2","SNAP_wave3")

# ----------- Function: run all four models for one wave/predictor combination-------------------
run_all_models <- function(split, wave, preds, suffix) {
  bind_rows(
    run_ablation(split, wave, preds,
                 paste("LR -", suffix), "glm"),
    run_ablation(split, wave, preds,
                 paste("CART -", suffix), "rpart",
                 tuneGrid = data.frame(cp = 0.01)),
    run_ablation(split, wave, preds,
                 paste("RF -", suffix), "rf",
                 tuneGrid = data.frame(mtry = floor(sqrt(length(preds))))),
    run_ablation_xgb(split, wave, preds,
                     paste("XGBoost -", suffix))
  )
}

# ----------- Run ablation ---------------------------
ablation_results <- bind_rows(

  # Wave 1 — demographics only (no SNAP history available)
  run_all_models(splits$w1, "Wave 1", demo_vars, "Demographics Only"),

  # Wave 2
  run_all_models(splits$w2, "Wave 2", demo_vars,             "Demographics Only"),
  run_all_models(splits$w2, "Wave 2", snap_w2,               "SNAP History Only"),
  run_all_models(splits$w2, "Wave 2", c(demo_vars, snap_w2), "Full Model"),

  # Wave 3
  run_all_models(splits$w3, "Wave 3", demo_vars,             "Demographics Only"),
  run_all_models(splits$w3, "Wave 3", snap_w3,               "SNAP History Only"),
  run_all_models(splits$w3, "Wave 3", c(demo_vars, snap_w3), "Full Model"),

  # Wave 4
  run_all_models(splits$w4, "Wave 4", demo_vars,             "Demographics Only"),
  run_all_models(splits$w4, "Wave 4", snap_w4,               "SNAP History Only"),
  run_all_models(splits$w4, "Wave 4", c(demo_vars, snap_w4), "Full Model")
)

print(ablation_results, n = 100)
write.csv(ablation_results, "ablation_results.csv", row.names = FALSE)

```

## Subgroup analyses [NOT IN THIS ARTICLE]

```{r}
# -------------- subgroup performance -----------------------------------------------
subgroup_performance <- function(test_data, preds, group_var, model_name) {
  test_data %>%
    mutate(.pred = preds) %>%
    group_by(across(all_of(group_var))) %>%
    summarise(
      model      = model_name,
      wave       = unique(wave),          
      subgroup   = group_var,
      group_val  = cur_group()[[1]],
      N          = n(),
      Accuracy   = mean(.pred == SNAP),
      Recall     = {
        tp <- sum(.pred == "Yes" & SNAP == "Yes")
        fn <- sum(.pred == "No"  & SNAP == "Yes")
        ifelse((tp + fn) == 0, NA, tp / (tp + fn))
      },
      Precision  = {
        tp <- sum(.pred == "Yes" & SNAP == "Yes")
        fp <- sum(.pred == "Yes" & SNAP == "No")
        ifelse((tp + fp) == 0, NA, tp / (tp + fp))
      },
      F1 = ifelse(is.na(Recall) | is.na(Precision) | (Recall + Precision) == 0,
                  NA, 2 * Recall * Precision / (Recall + Precision)),
      .groups = "drop"
    )
}

# ------------ Fit & predict for one wave -----------------------
run_wave_subgroups <- function(wave_splits, wave_label, subgroup_vars) {
  
  train_d <- wave_splits$train
  test_d  <- wave_splits$test
  ctrl    <- trainControl(method = "none")
  
  # --- Logistic Regression ---
  lr_fit   <- train(SNAP ~ ., data = train_d, method = "glm",
                    family = "binomial", trControl = ctrl)
  lr_preds <- predict(lr_fit, test_d)
  
  # --- CART ---
  cart_fit   <- train(SNAP ~ ., data = train_d, method = "rpart",
                      trControl = ctrl,
                      tuneGrid = data.frame(cp = 0.01))
  cart_preds <- predict(cart_fit, test_d)
  
  # --- Random Forest ---
  rf_fit   <- train(SNAP ~ ., data = train_d, method = "rf",
                    trControl = ctrl,
                    tuneGrid = data.frame(mtry = floor(sqrt(ncol(train_d) - 1))),
                    ntree = 100)
  rf_preds <- predict(rf_fit, test_d)
  
  # --- XGBoost ---
  X_tr    <- model.matrix(SNAP ~ . - 1, data = train_d)
  X_te    <- model.matrix(SNAP ~ . - 1, data = test_d)
  y_tr    <- as.numeric(train_d$SNAP == "Yes")
  xgb_fit <- xgb.train(
    params   = list(objective = "binary:logistic", eta = 0.1,
                    max_depth = 6, nthread = 4),
    data     = xgb.DMatrix(X_tr, label = y_tr),
    nrounds  = 100,
    verbose  = 0
  )
  xgb_preds <- factor(
    ifelse(predict(xgb_fit, xgb.DMatrix(X_te)) >= 0.5, "Yes", "No"),
    levels = c("No", "Yes")
  )
  
  # --- Subgroup results for this wave ---
  test_d$wave <- wave_label   # tag the wave so it surfaces in output
  
  bind_rows(
    map_dfr(subgroup_vars, ~subgroup_performance(test_d, lr_preds,   .x, "Logistic Regression")),
    map_dfr(subgroup_vars, ~subgroup_performance(test_d, cart_preds, .x, "CART")),
    map_dfr(subgroup_vars, ~subgroup_performance(test_d, rf_preds,   .x, "Random Forest")),
    map_dfr(subgroup_vars, ~subgroup_performance(test_d, xgb_preds,  .x, "XGBoost"))
  )
}

# ------------Run across all four waves --------------------------
subgroup_vars <- c("race", "education", "employment", "income")

all_wave_subgroup_results <- bind_rows(
  run_wave_subgroups(splits$w1, "Wave 1", subgroup_vars),
  run_wave_subgroups(splits$w2, "Wave 2", subgroup_vars),
  run_wave_subgroups(splits$w3, "Wave 3", subgroup_vars),
  run_wave_subgroups(splits$w4, "Wave 4", subgroup_vars)
)

# ── Inspect & save ────────────────────────────────────────────────────────────
print(all_wave_subgroup_results, n = 200)
write.csv(all_wave_subgroup_results, "subgroup_results_all_waves.csv", row.names = FALSE)
```


## Attrition analysis (whether retaining the part of the sample introduces systematic bias?)


```{r}
# ------------ 1. BUILD THE RETAINED/ATTRITED FLAG ------------------------
baseline_w1 <- w1 %>%
  filter(imputed_snap == FALSE) %>%
  select(ID_nozero, marital.status, sex, hispanic.origin,
         race, education, citizenship, employment, state, age, income,
         snap)   # ── NEW: include Wave 1 SNAP outcome

ids_w2 <- w2 %>% filter(imputed_snap == FALSE) %>% pull(ID_nozero)
ids_w3 <- w3 %>% filter(imputed_snap == FALSE) %>% pull(ID_nozero)
ids_w4 <- w4 %>% filter(imputed_snap == FALSE) %>% pull(ID_nozero)

ids_retained <- Reduce(intersect, list(
  baseline_w1$ID_nozero, ids_w2, ids_w3, ids_w4
))

attrition_df <- baseline_w1 %>%
  mutate(
    status = if_else(ID_nozero %in% ids_retained, "Retained", "Attrited"),
    status = factor(status, levels = c("Retained", "Attrited"))
  )

cat("Wave 1 baseline N        :", nrow(attrition_df), "\n")
cat("Retained across all waves:", sum(attrition_df$status == "Retained"), "\n")
cat("Attrited (≥1 wave missing):", sum(attrition_df$status == "Attrited"), "\n")
cat("Attrition rate            :",
    round(mean(attrition_df$status == "Attrited") * 100, 1), "%\n")


# ── 2. SMD HELPER FUNCTIONS ──────────────────────────────────────────────────  # ── NEW
smd_continuous <- function(x, group) {
  # Cohen's d using pooled SD
  vals   <- split(as.numeric(as.character(x)), group)
  m_diff <- mean(vals[["Retained"]], na.rm = TRUE) -
            mean(vals[["Attrited"]],  na.rm = TRUE)
  pooled_sd <- sqrt((var(vals[["Retained"]], na.rm = TRUE) +
                     var(vals[["Attrited"]],  na.rm = TRUE)) / 2)
  round(abs(m_diff / pooled_sd), 3)
}

smd_categorical <- function(var_vec, group_vec) {
  # Average of absolute SMDs across all categories (for multi-level vars)
  cats <- levels(factor(var_vec))
  smds <- sapply(cats, function(cat) {
    p1 <- mean(var_vec[group_vec == "Retained"] == cat, na.rm = TRUE)
    p2 <- mean(var_vec[group_vec == "Attrited"]  == cat, na.rm = TRUE)
    p_bar <- (p1 + p2) / 2
    denom <- sqrt(p_bar * (1 - p_bar))
    if (denom == 0) return(0)
    abs(p1 - p2) / denom
  })
  round(mean(smds), 3)
}


# ------------ 3. DEMOGRAPHIC COMPARISON TABLE ------------------------------------------
demo_vars <- c("snap",                                           # ── NEW
               "sex", "race", "hispanic.origin", "education",
               "marital.status", "employment", "citizenship", "income")

# Age (continuous)
age_summary <- attrition_df %>%
  group_by(status) %>%
  summarise(
    mean_age = round(mean(as.numeric(as.character(age)), na.rm = TRUE), 1),
    sd_age   = round(sd(as.numeric(as.character(age)),  na.rm = TRUE), 1),
    .groups  = "drop"
  ) %>%
  mutate(label = paste0(mean_age, " (", sd_age, ")")) %>%
  select(status, label) %>%
  pivot_wider(names_from = status, values_from = label) %>%
  mutate(
    Variable = "Age",
    Category = "Mean (SD)",
    p_value  = {
      age_num <- as.numeric(as.character(attrition_df$age))
      t.test(age_num ~ attrition_df$status)$p.value
    },
    SMD = smd_continuous(attrition_df$age, attrition_df$status)  # ── NEW
  )

# Categorical variables
make_cat_table <- function(var) {
  tbl <- attrition_df %>%
    group_by(status, .data[[var]]) %>%
    tally() %>%
    group_by(status) %>%
    mutate(pct = round(n / sum(n) * 100, 1)) %>%
    mutate(cell = paste0(n, " (", pct, "%)")) %>%
    select(status, category = all_of(var), cell) %>%
    pivot_wider(names_from = status, values_from = cell,
                values_fill = "0 (0.0%)") %>%
    mutate(Variable = var)

  chi_p <- tryCatch(
    chisq.test(table(attrition_df$status, attrition_df[[var]]))$p.value,
    error = function(e) NA_real_
  )

  smd_val <- smd_categorical(attrition_df[[var]], attrition_df$status)  # ── NEW

  tbl %>%
    mutate(
      p_value = if_else(row_number() == 1, chi_p, NA_real_),
      SMD     = if_else(row_number() == 1, smd_val, NA_real_)           # ── NEW
    )
}

cat_tables <- bind_rows(lapply(demo_vars, make_cat_table)) %>%
  rename(Category = category)

# Format p-values
fmt_p <- function(p) {
  case_when(
    is.na(p)   ~ "",
    p < 0.001  ~ "<0.001",
    p < 0.01   ~ format(round(p, 3), nsmall = 3),
    TRUE       ~ format(round(p, 2), nsmall = 2)
  )
}

# Combine
attrition_table <- bind_rows(
  cat_tables %>% select(Variable, Category, Retained, Attrited, p_value, SMD),
  age_summary %>% select(Variable, Category, Retained, Attrited, p_value, SMD)
) %>%
  arrange(Variable) %>%
  mutate(p_value = fmt_p(p_value),
         SMD     = if_else(is.na(SMD), "", as.character(SMD)))  # ── NEW

# ── 4. OUTPUT ─────────────────────────────────────────────────────────────────
attrition_table %>%
  select(Variable, Category, Retained, Attrited,
         `p-value` = p_value, SMD) %>%           # ── NEW: SMD column added
  kable(
    format   = "latex",
    booktabs = TRUE,
    caption  = "Demographic characteristics of retained versus attrited
respondents. Retained respondents were observed across all four SIPP waves
with non-imputed SNAP outcomes; attrited respondents were present at
Wave~1 but missing from at least one subsequent wave. Percentages are
column proportions within each group. $p$-values are from chi-square tests
(categorical variables) and two-sample $t$-tests (age). SMD denotes the
standardized mean difference; values $\\geq 0.10$ indicate meaningful
imbalance. Attrited respondents are systematically younger, less educated,
more likely to be never married, and more likely to belong to racial and
ethnic minority groups, consistent with positive selection of the retained
sample on socioeconomic stability. See Appendix~\\ref{app:attrition} for
implications for generalizability.",
    label    = "tab:attrition"
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  collapse_rows(columns = 1, latex_hline = "major")

write.csv(attrition_table, "attrition_table2_appendix.csv", row.names = FALSE)

# ── 3. WAVE-BY-WAVE SAMPLE FLOW TABLE ────────────────────────────────────────
# Shows N retained at each cumulative threshold — useful for the paper body

wave_flow <- tibble(
  Wave = c("Wave 1 (baseline)", "Through Wave 2",
           "Through Wave 3", "Through Wave 4"),
  N_nonimputed = c(
    nrow(baseline_w1),
    length(intersect(baseline_w1$ID_nozero, ids_w2)),
    length(Reduce(intersect, list(baseline_w1$ID_nozero, ids_w2, ids_w3))),
    length(ids_retained)
  )
) %>%
  mutate(
    Pct_of_baseline = round(N_nonimputed / nrow(baseline_w1) * 100, 1),
    Wave_to_wave_loss = N_nonimputed - lag(N_nonimputed, default = N_nonimputed[1]),
    Wave_to_wave_loss = if_else(Wave_to_wave_loss == 0, NA_integer_, -Wave_to_wave_loss)
  )

print(wave_flow)

wave_flow %>%
  rename(
    `Cumulative N` = N_nonimputed,
    `% of Wave 1`  = Pct_of_baseline,
    `N lost`       = Wave_to_wave_loss
  ) %>%
  kable(
    format   = "latex",
    booktabs = TRUE,
    caption  = "Cumulative sample retention across waves. Only respondents with self-reported (non-imputed) SNAP outcomes at all prior waves are retained.",
    label    = "tab:wave_flow"
  ) %>%
  kable_styling(latex_options = "hold_position")


# ----------- 4. ATTRITION PROFILE FIGURE --------------------------
# Grouped bar chart: % distribution of key demographics by retained/attrited
# Keeps the figure compact — use the 4 most policy-relevant variables

plot_vars <- c("sex", "race", "education", "marital.status", "employment")

plot_df <- bind_rows(lapply(plot_vars, function(var) {
  attrition_df %>%
    group_by(status, cat = .data[[var]]) %>%
    tally() %>%
    group_by(status) %>%
    mutate(pct = n / sum(n) * 100) %>%
    mutate(variable = var)
}))

ggplot(plot_df, aes(x = cat, y = pct, fill = status)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~ variable, scales = "free_x", ncol = 3) +
  scale_fill_manual(
    values = c("Retained" = "#2166ac", "Attrited" = "#d6604d"),
    name   = NULL
  ) +
  labs(
    x     = NULL,
    y     = "Percentage within group (%)",
    title = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 8),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

ggsave("appendix_attrition_profile.pdf", width = 7, height = 6)
cat("Figure saved: appendix_attrition_profile.pdf\n")
```
